---
title: Probabilistic Changepoint Modeling
subtitle: An example report for Pittsburgh useR Group
author: 
  - Mikhail Popov
date: "`r sub('^0(\\d)', '\\1', format(Sys.Date(), '%d %B %Y'))`"
abstract: |
  Our imaginary store receives some number of customers per day. We start an advertising campaign which takes some time to have a full effect on the rate of customers we receive daily, and then has a long-term effect after we have stopped it. Using probabilistic programming languages (PPLs), we can specify a Bayesian model and infer the hidden rates.
output:
  wmfpar::pdf_report:
    cite_r_packages:
      - kableExtra
      - rstan
      - bridgesampling
      - purrr
      - dplyr
      - tidyr
      - ggplot2
      - import
      - here
nocite: "@*"
---

```{r setup, include=FALSE}
library(knitr)
library(here)
opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = "png", dpi = 600, digits = 3
)
options(knitr.table.format = "latex")
```
```{r pkgs}
library(magrittr) # piping
import::from(dplyr, data_frame)
library(kableExtra) # table formatting
library(ggplot2) # visualization
library(rstan) # modelling & inference
library(bridgesampling) # model comparison
# RStan Config:
# options(mc.cores = parallel::detectCores())
options(mc.cores = 1)
rstan_options(auto_write = TRUE)
```

# Data simulation

Suppose we are operating a store which receives a random number of customers per day, and specifically it's random according to the Poisson distribution with rate $\lambda_1$. Then we start an advertisement campaign -- which takes a few days to come to full effect -- that changes the rate to $\lambda_2$. When we stop the ad, it gradually loses effect but the campaign has had a long-lasting effect (e.g. we've gained new customers who come back regularly), which means our store now receives customers at rate $\lambda_3$.

```{r parameters}
lambda_1 <- 300
lambda_2 <- 400
lambda_3 <- 350
transition_period_1 <- 7 # days
transition_period_2 <- 21 # days
total_days <- 180
change_point_1 <- 60
change_point_2 <- 90
```

```{r param_tables}
data_frame(
  parameter = c(
    "$\\lambda_1$ (old normal)", "$\\lambda_2$ (temporary)", "$\\lambda_3$ (new normal)",
    "$N$ (total days)", "$T_1$ (ad start)", "$T_2$ (ad stop)",
    "$d_1$ (time for full effect)", "$d_2$ (time to new normal)"
  ),
  value = c(
    lambda_1, lambda_2, lambda_3,
    total_days, change_point_1, change_point_2,
    transition_period_1, transition_period_2
  )
) %>%
  kable(
    escape = FALSE, booktabs = TRUE,
    caption = "Simulation parameters"
  ) %>%
  kable_styling(latex_options = "hold_position") %>%
  group_rows(index = c("Rates (unknown)" = 3, "Other parameters" = 5))
```

In the simulation, we have two weights -- $w_1$ and $w_2$ -- which sum to 1 and produce gradual (albeit not smooth) transitions between the different rates. The change is a slope in this toy scenario, but it can also be a more interesting transition like a sigmoid function such as the [Gompertz curve](https://en.wikipedia.org/wiki/Gompertz_function).

```{r simulation}
weight_func <- function(i, length_out) {
  weight_1 <- seq(1, 0, length.out = length_out + 2)[2:(length_out + 1)]
  weight_2 <- 1 - weight_1
  return(c(weight_1[i], weight_2[i]))
}
rcustomer <- function(t, T1, T2, d1, d2, l1, l2, l3, wf) {
  if (t <= T1) {
    l <- l1
  } else if (t > T2 + d2) {
    l <- l3
  } else {
    if (t > T1 && t <= T1 + d1) {
      i <- t - T1
      w <- wf(i, d1)
      l <- w[1] * l1 + w[2] * l2
    } else if (t > T2 && t <= T2 + d2) {
      i <- t - T2
      w <- wf(i, d2)
      l <- w[1] * l2 + w[2] * l3
    } else {
      l <- l2
    }
  }
  n <- rpois(1, l)
  return(data_frame(day = t, rate = l, customers = n))
}
```

```{r customers}
set.seed(42)
customers <- purrr::map_dfr(
  1:total_days, rcustomer, wf = weight_func,
  T1 = change_point_1, T2 = change_point_2,
  d1 = transition_period_1, d2 = transition_period_2,
  l1 = lambda_1, l2 = lambda_2, l3 = lambda_3
)
```

```{r plot, fig.width=8, fig.height=4, dpi=300}
annotations <- data_frame(
  day = c(change_point_1, change_point_2),
  event = c("Start", "Stop")
)
ggplot(customers, aes(x = day, y = customers)) +
  geom_vline(
    aes(xintercept = day), data = annotations,
    linetype = "dashed", color = "blue"
  ) +
  geom_line() +
  geom_line(aes(y = rate), color = "blue", size = 1.1) +
  geom_label(
    aes(label = event, y = 250), data = annotations,
    color = "blue"
  ) +
  theme_minimal(14, "Source Sans Pro") +
  labs(
    x = "Day", y = "Customers per day",
    title = "Daily customers before, during, and after an advertising campaign"
  )
```

# Modelling

We'll take a look at three similar models -- $\mathcal{M}_1, \mathcal{M}_2, \mathcal{M}_3$ -- which model the daily counts of customers $y_t$ as a [Poisson](https://en.wikipedia.org/wiki/Poisson_distribution)-distributed random variable with time-varying rate $\lambda(t)$, up to a maximum of $t = N$ days of data: $y_t~\sim~\text{Poisson}(\lambda(t)),~t = 1,\ldots,N.$

The ad campaign starts on $T_1$ and stops on $T_2$. We are interested in inferring $\lambda_1$ (the rate before $T_1$), $\lambda_2$ (the rate between $T_1$ and $T_2$), and $\lambda_3$ (the rate after $T_2$). We specify $\lambda~\sim~\mathcal{N}(300, 100)$ and let Stan implicitly assign a default prior to $\beta_1$ and $\beta_2$.

## Model 1

In the simpler model $\mathcal{M}_1$, we ignore the obvious transition periods and model the switch between the rates as immediate:

$$
\lambda(t) = \begin{cases}
  \lambda_1, & \text{if } t \leq T_1, \\
  \lambda_2, & \text{if } T_1 < t \leq T_2, \\
  \lambda_3, & \text{if } t > T_2.
\end{cases}
$$

## Model 2

In the slightly more complex model $\mathcal{M}_2$, we include the two gradual changes as slopes over $d_1$ and $d_2$ days after $T_1$ and $T_2$, respectively. We formalize this as follows:

$$
\lambda(t) = \begin{cases}
  \lambda_1, & \text{if } t \leq T_1, \\
  \lambda_2, & \text{if } T_1 + d_1 \leq t \leq T_2, \\
  \lambda_3, & \text{if } t \geq T_2 + d_2, \\
  \lambda_1 + \beta_1 (t - T_1), & \text{if } T_1 < t \leq T_1 + d_1, \\
  \lambda_2 + \beta_2 (t - T_2), & \text{if } T_2 < t \leq T_2 + d_2.
\end{cases}
$$

**Note**: in this toy scenario we know $d_1$ and $d_2$, but we can also make them hidden parameters to infer from the data. We could formalize our intuition about their values by assigning them priors $\mathcal{N}(7, 2)$ and $\mathcal{N}(14, 4)$, respectively.

\clearpage

# Model comparison

```{r models, message=FALSE}
models <- list(
  model_1 = stan_model("models/model_1.stan", "model_1"),
  model_2 = stan_model("models/model_2.stan", "model_2")
)
```

```{r mcmc_draw, cache=TRUE}
data <- list(
  N = total_days, y = customers$customers,
  T = c(change_point_1, change_point_2),
  d = c(transition_period_1, transition_period_2)
)

draws <- purrr::map(models, sampling, data = data, refresh = 0)
```

```{r bridge_samples, cache=TRUE, dependson='mcmc_draws'}
bridge_samples <- purrr::map(draws, bridge_sampler, silent = TRUE)
```

```{r helpers}
extract_ci = function(draws, ...) {
  extracted_ci <- draws %>%
    broom::tidyMCMC(pars = "lambda", conf.int = TRUE, ...) %>%
    dplyr::rename(est = estimate) %>%
    dplyr::mutate(est = sprintf("%.1f (%.1f--%.1f)", est, conf.low, conf.high)) %>%
    dplyr::select(term, est) %>%
    tidyr::spread(term, est)
  return(extracted_ci)
}
```

```{r model_comparison}
data_frame(
  model = names(bridge_samples),
  post_prob = post_prob(purrr::map_dbl(bridge_samples, logml))
) %>%
  dplyr::left_join(purrr::map_dfr(draws, extract_ci, .id = "model"), by = "model") %>%
  dplyr::mutate(
    post_prob = sprintf("%.4f", post_prob),
    model = sub("model_([1-3])", "$\\\\mathcal{M}_\\1$", model),
  ) %>%
  kable(
    col.names = c("Model", "$\\text{Pr}(\\mathcal{M}|\\mathcal{D})$",
                  sprintf("$\\lambda_%i$ (%.0f)", 1:3, c(lambda_1, lambda_2, lambda_3))),
    booktabs = TRUE, escape = FALSE, align = c("l", "r", "c", "c", "c"),
    caption = "Posterior probabilities of the three models given data and the estimated rates, calculated using the bridgesampling package."
  ) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  add_header_above(header = c(" " = 2, "Estimate (95% Credible Interval)" = 3))
```

The [Bayes factor](https://en.wikipedia.org/wiki/Bayes_factor) (BF) can be used to decide between these two competing models -- $\mathcal{M}_1$ (instant changes between rates) and $\mathcal{M}_2$ (gradual changes between rates) -- by quantifying how much more likely the data $\mathcal{D}$ is under $\mathcal{M}_1$ vs $\mathcal{M}_2$:

$$
\text{BF}_{12} = \frac{p(\mathcal{D}|\mathcal{M}_1)}{p(\mathcal{D}|\mathcal{M}_2)},
$$

Using **bridgesampling** [@bridgesampling] allows us to calculate marginal likelihoods -- $p(\mathcal{D}|\mathcal{M})$ -- of Stan models really easily and therefore compute the BF (see `?bf`), which -- for $\mathcal{M}_1$ compared to $\mathcal{M}_2$, for example -- comes out to be `r round(bf(bridge_samples$model_1, bridge_samples$model_2)$bf, 2)`, meaning there is no evidence for choosing model 1 over model 2.

## Inference results

```{r results}
differences <- c(
  lambda_2 - lambda_1,
  lambda_3 - lambda_2,
  lambda_3 - lambda_1
)
draws$model_2 %>%
  broom::tidyMCMC(
    pars = c("lambda", "diff21", "diff32", "diff31"),
    estimate.method = "median", conf.int = TRUE, conf.method = "HPDinterval"
  ) %>%
  dplyr::mutate(
    term = c(
      paste0("$\\lambda_", 1:3, "$"),
      "$\\lambda_2 - \\lambda_1$",
      "$\\lambda_3 - \\lambda_2$",
      "$\\lambda_3 - \\lambda_1$"
    ),
    truth = c(
      lambda_1, lambda_2, lambda_3,
      differences
    ),
    conf.int = sprintf("(%.1f, %.1f)", conf.low, conf.high)
  ) %>%
  dplyr::select(term, truth, estimate, conf.int) %>%
  kable(
    escape = FALSE, booktabs = TRUE, digits = 1,
    align = c("l", "r", "r", "c"),
    col.names = c("Parameter", "Truth", "Point Estimate", "95\\% Credible Interval"),
    caption = "Inference using $\\mathcal{M}_2$ (gradual changes between rates)."
  ) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  group_rows(index = c("Rates" = 3, "Differences" = 3))
```

Table 3 shows the inferred differences of rates and we can see that our imaginary advertisement had a positive, statistically significant impact on how many imaginary customers our imaginary store receives per day on average, both during the campaign and long after.

\newpage

# References

\footnotesize
